Treatment planning for HDR Brachytherapy

### 2015-03-28 Paper Seminar/isim_paper.tex
Removed duplicate bibliography
added basic title page and VOI OAR PTV definitions in model description
Included classification file

\begin{document}
\tableofcontents

\section{Model description}
HDR brachytherapy consists of several radioactive sources being placed inside the patient's body by needles. These sources will eradiate into the surrounding tissues, treating tumor cells and other critical volumes. Therefore, it relies heavily on optimal algorithms to determine the dwell times for certain, predefined goals. \\
The body is represented by an equidistant grid of voxels, defining the single body cell type. The seeds are randomly positioned within the tumor voxels. Hence, the dosis in a voxel is calculated as a sum of each seed by evaluating the dose-function up to a fixed limit, where the influence gets negligible.

@@

@@

For our problem we used CPLEX, an implementation of the simplex algorithm by IBM[?]. In simplex the feasible region is considered a polyhedron. The optimal solution must be in one of the vertices of the polyhedron. Simplex finds a start vertex and approaches the optimal solution by going on an edge from vertex to vertex, until there is no better solution.
The constraints allow us a good modelling of the tumor, since we are able to define a constraint with certain limits for each body type.

\paragraph{Basic Funtionality}
The basic functionality is depicted in algorithm \ref{alg:cplex}. This implementation checks every voxel of the body, if it is a tumor voxel or not. In case of a tumor voxel the influence of the radiation of every seed will be summed up, building a greater-equal constraint. For body voxel the influence of the radiation builds a term for the objective function. 
\begin{algorithm}
\label{alg:cplex}
 \For{x=0; x $\textless$ body.xDim; x++}{
 	\For{y=0; y $\textless$ body.yDim; y++}{
 		\For{z=0; z $\textless$ body.zDim; z++}{
			\If{$body[x][y][z].getBodyType() == tumorType$} {
				\For{i=0; i $\textless$ numberOfSeeds; i++}{
					$dosepart[i] \gets cplex.prod(dose(x,y,z,i),time[i])$	
				}
				$cplex.addGe(cplex.sum(dosepart),goalDosis(x,y,z));$
				}
			\Else{\For{i=0; i $\textless$ numberOfSeeds; i++}{
					$objective.addTerm(dose(x,y,z,i),time[i]);$	
				}

@@

@@

\begin{thebibliography}{9}
\bibitem{schlaefer}
	Schlaefer, A. and Schweikard, A. Stepwise multi-criteria optimization for robotic radiosurgery. Med. Phys. 2008.

\end{thebibliography}
\end{document}

>>>>

\begin{document}

\title{Treatment planning for HDR Brachytherapy}
\date{March 30, 2015}
\author{S.~Elm \and
T.~Karthe \and
M.~E.~Schroeder \and
D.~Golebiewski \and
L.~Mordhorst}

\maketitle
\newpage

\tableofcontents

\section{Model description}
HDR brachytherapy consists of several radioactive sources being placed inside the patient's body by needles. These sources will eradiate into the surrounding tissues, treating tumor cells and other critical volumes. Therefore, it relies heavily on optimal algorithms to determine the dwell times for certain, predefined goals. \\
The body is represented by an equidistant grid of voxels, defining the single body cell type. The seeds are randomly positioned within the tumor voxels. Hence, the dosis in a voxel is calculated as a sum of each seed by evaluating the dose-function up to a fixed limit, where the influence gets negligible. \\
In the model the tumor is called the planning target volume (PTV) and the surrounding organs are called organs at risk (OAR). The different body types might also be referred as volume of interest (VOI).

@@

\section{Classification}
\input{classification}

@@

For our problem we used CPLEX, an implementation of the simplex algorithm by IBM[?]. In simplex the feasible region is considered a polyhedron. The optimal solution must be in one of the vertices of the polyhedron. Simplex finds a start vertex and approaches the optimal solution by moving from vertex to vertex on an edge, until there is no better solution.
The constraints allow us a good modelling of the tumor, since we are able to define a constraint with certain limits for each body type.

\subsubsection{Basic Funtionality}

\begin{algorithm}[H]
\label{alg:cplex}
 \For{x=0; x $\textless$ body.xDim; x++}{
 	\For{y=0; y $\textless$ body.yDim; y++}{
 		\For{z=0; z $\textless$ body.zDim; z++}{
			\If{$body[x][y][z].getBodyType() == tumorType$} {
				\For{i=0; i $\textless$ numberOfSeeds; i++}{
					$dosepart[i] \gets cplex.prod(dose(x,y,z,i),time[i])$	
				}
				$cplex.addGe(cplex.sum(dosepart),goalDosis(x,y,z));$
				}
			\Else{\For{i=0; i $\textless$ numberOfSeeds; i++}{
					$objective.addTerm(dose(x,y,z,i),time[i]);$	
				}
				}

@@

The basic functionality is depicted in algorithm \ref{alg:cplex}. This implementation checks every voxel of the body, if it is a tumor voxel or not. In case of a tumor voxel the influence of the radiation of every seed will be summed up, building a greater-equal constraint. For body voxels the influence of the radiation builds a term for the objective function.

 \subsubsection{Results}
 The constraints restrict the optimization so that the level of radiation in the PTV does not fall under a fixed value. The contraints can be relaxed to lower the level of radiation in the surrounding OAR at the cost of lowering the radiation at the border of the PTV.

 As shown in table \ref{table:LP_results50} and \ref{table:LP_results100}, the relaxation of the PTV increases the conformality, while letting the coverage of the PTV drop. 
		\begin{table}[h]
			\centering		
		 	\caption{Treatment quality with 50 seeds and relaxation}
		 	\label{table:LP_results50}
			\begin{tabular}{C{0.35\textwidth}|C{0.11\textwidth}|C{0.1\textwidth}}
			\textbf{PTV lower bound [Gy]} 	& \textbf{CO} 	& \textbf{CI}\\	\hline
				32.0 	& 0.999		& 1.23\\	\hline
				31.0 	& 0.988 	& 1.20\\	\hline
				30.0 	& 0.953		& 1.17\\	\hline		
				29.0 	& 0.899		& 1.14\\
			\end{tabular}
		\end{table}

Using more initial seeds, the algorithm can achieve higher conformality without losing coverage.

				\begin{table}[h]
			\centering		
		 	\caption{Treatment quality with 100 seeds and relaxation}
		 	\label{table:LP_results100}
			\begin{tabular}{C{0.35\textwidth}|C{0.11\textwidth}|C{0.1\textwidth}}
			\textbf{PTV lower bound [Gy]} 	& \textbf{CO} 	& \textbf{CI}\\	\hline
				32.0 	& 0.999		& 1.16\\ \hline
				31.0 	& 0.981 	& 1.13\\ \hline
				30.0 	& 0.931		& 1.10\\ \hline	
				29.0 	& 0.851		& 1.08\\
			\end{tabular}
		\end{table}

@@

\end{document}
